---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
Information about my environment. Have a 'minor issue' that knitr in rstudio does not use the same library path.

Well i have a few hours before due date - it could be worse...

```{r}
 .libPaths()
```
To install the required libraries ran the following.
```{r, cache=TRUE, echo=TRUE}
#install.packages("lubridate")
#install.packages("dplyr")
#install.packages("ggplot2")
#install.packages("gridExtra")
```

```{r echo=TRUE, message=FALSE}
library(lubridate)
library(dplyr)
library(ggplot2)
library(gridExtra)
```
### Loading and preprocessing the data

First step is to read the data. This is a local copy of the data available at https://github.com/rdpeng/RepData_PeerAssessment1.

Read in data using read.delim() which is same family as read.csv(). The data frame **df.activity** is used for the raw data. Later **df.activity.padded** will be used for the data where the NA's have been replaced.


```{r}
dir<-'C:/_store/@stats/data-course/3.-reproducable-research-Oct2014/assignment_1'
setwd(dir)

file<- paste(dir,'/activity.csv', sep='')
df.activity<-read.delim(file, header = TRUE, sep = ",", quote = "\"", stringsAsFactors =FALSE)

```

For analysis i need to add some date and time information to this data frame
```{r}

# minor tweak with data to deal with date  #########################
# inital date is a chr column
colnames(df.activity)<-c('steps','datechr','interval')

#to make a datetime column
#nb. dates are strings in "yyyy-mm-dd"
df.activity$date<-ymd(df.activity$datechr)

###make a datetime column #################################

df.activity$hhmm<-sprintf("%04d", df.activity$interval)
df.activity$datetime<-paste(df.activity$datechr, df.activity$hhmm, ' ')
df.activity$datetime<-ymd_hm(df.activity$datetime)

#to get just the time during the day - will use decimal hours - in plots

df.activity$time<- hour(df.activity$datetime)*60+minute(df.activity$datetime)
#fractional hours
df.activity$hrs<- hour(df.activity$datetime)+minute(df.activity$datetime)/60

#a number of tasks require separation of weekday and weekend

#determine if day is in the weekend or is a weekday
#wday() returns the day of the week as a decimal number (01-07, Sunday is 1).
df.activity$wday<-wday(df.activity$date)

#setup the weekday/weekend factor as a column
df.activity$weekend <- as.factor(
  ifelse(df.activity$wday == 1 | df.activity$wday==7 , 'weekend', 'weekday')
)

## and summary info
str(df.activity)
summary(df.activity)

```

### What is mean total number of steps taken per day?



```{r}

```

### What is the average daily activity pattern?


```{r}

```


### Imputing missing values


```{r}

```


### Are there differences in activity patterns between weekdays and weekends?


```{r}

```

