---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
Information about my environment. Have a 'minor issue' that knitr in rstudio does not use the same library path and so will not run the Rmd script.

Well i have a few hours before due date - it could be worse...

```{r}
 .libPaths()
```
To install the required libraries ran the following.
```{r, cache=TRUE, echo=TRUE}
#install.packages("lubridate")
#install.packages("dplyr")
#install.packages("ggplot2")
#install.packages("gridExtra")
```

```{r echo=TRUE, message=FALSE}
library(lubridate)
library(dplyr)
library(ggplot2)
library(gridExtra)
```
### Loading and preprocessing the data

First step is to read the data. This is a local copy of the data available at https://github.com/rdpeng/RepData_PeerAssessment1.

Read in data using read.delim() which is same family as read.csv(). The data frame **df.activity** is used for the raw data. Later **df.activity.padded** will be used for the data where the NA's have been replaced.


```{r}
dir<-'C:/_store/@stats/data-course/3.-reproducable-research-Oct2014/assignment_1'
setwd(dir)

file<- paste(dir,'/activity.csv', sep='')
df.activity<-read.delim(file, header = TRUE, sep = ",", quote = "\"", stringsAsFactors =FALSE)

```

For analysis i need to add some date and time information to this data frame
```{r}

# minor tweak with data to deal with date  #########################
# inital date is a chr column
colnames(df.activity)<-c('steps','datechr','interval')

#to make a datetime column
#nb. dates are strings in "yyyy-mm-dd"
df.activity$date<-ymd(df.activity$datechr)

###make a datetime column #################################

df.activity$hhmm<-sprintf("%04d", df.activity$interval)
df.activity$datetime<-paste(df.activity$datechr, df.activity$hhmm, ' ')
df.activity$datetime<-ymd_hm(df.activity$datetime)

#to get just the time during the day - will use decimal hours - in plots

df.activity$time<- hour(df.activity$datetime)*60+minute(df.activity$datetime)
#fractional hours
df.activity$hrs<- hour(df.activity$datetime)+minute(df.activity$datetime)/60

#a number of tasks require separation of weekday and weekend

#determine if day is in the weekend or is a weekday
#wday() returns the day of the week as a decimal number (01-07, Sunday is 1).
df.activity$wday<-wday(df.activity$date)

#setup the weekday/weekend factor as a column
df.activity$weekend <- as.factor(
  ifelse(df.activity$wday == 1 | df.activity$wday==7 , 'weekend', 'weekday')
)

## and summary info
str(df.activity)
summary(df.activity)

```
### Imputing missing values.

the NA values are replaced by the following algorith.

1. calculating the mean of the interval across the 2 months taking into account the weekend as a factor. 
2. these are joined into the df.activity dataframe so that available as a 'mean_steps'.
3. where the row has a steps of NA using the mean_steps is used to replace the NA

This analysis, using dplyr, is done here, early so that the following plots can compare the raw and padded data.

An insight had during experimentation was that this replacment does shift the histogram as there are less zero activity days but does not shift the mean activity - as just added back the mean activity.

```{r }

############################################################
### clean up data by replacing the steps=NA with the mean number of steps taken in that interval.
### as the weekend and weekday will influence this mean
### use mean of weekday interval and mean of weekend interval

### firstly get the interval_mean_steps_w where there are means for 
# weekdays and weekend intervals

interval_steps_ww <- 
  df.activity  %>%
  group_by( weekend, interval)  %>%
  summarise(
    mean_steps = mean(steps, na.rm = TRUE)
  )

# add on a joining column 'interval_weekend' so that dplyr join() can be used (see below)
interval_steps_ww <- 
  interval_steps_ww  %>%
  mutate ( interval_weekend = paste (weekend,interval))

# add that same column to the df.activity dataframe 
 
df.activity<-
df.activity %>%
  mutate ( interval_weekend = paste (weekend,interval))

# join the dataframes back to df.activity so have mean_steps available if steps is NA
# this bvings the intervals mean in so that it can be used to replace the NA where appropriate
df.activity<-left_join(df.activity, interval_steps_ww, by=c("interval_weekend")) 

#clean up from the join 
df.activity <-
  df.activity %>%
  select( steps, datechr, interval=interval.x, date, hhmm, datetime, time, hrs, wday, weekend=weekend.x, mean_steps)
  
#create another dataframe replacing all the NA's with the interval mean of the weekday or weekend 
# df.activity.padded has the NA replaced otherwise the same as df.activity

df.activity.padded<- 
  df.activity %>%
  mutate(
    steps =
      ifelse(
        is.na( steps), 
        mean_steps, #an alternative that does make a visible difference is to set the NA to zero 
        steps
      )
  )
```

```{r}
## and summary info
str(df.activity)
summary(df.activity)

str(df.activity.padded)
summary(df.activity.padded)

```

```

### What is mean total number of steps taken per day?

Use a function histo_mean_median() to generate a list containing the plot and the mean and median.
It will later be called for both the raw and padded data.

```{r}
# require a histogram of the steps per day and mean and median
# need this twice for different data so do it in a function

histo_mean_median <- function (df, title) {
  # get the steps per day
  df.perday <- 
    df %>%
    group_by( date)  %>%
    summarise(
      steps_per_day = sum(steps, na.rm = TRUE)
    )
  #plot it as a histogram - force scales to be same for comparison
  p<-ggplot(df.perday, aes(x=steps_per_day)) +
          geom_histogram(binwidth = 250) + 
          labs( title=title ,
                x = 'steps per day',
                y = 'number of days')+
          xlim(0,25000) +
          scale_y_continuous(breaks=0:10)
  # what where the mean and medium of step per day
  mean_median <- df.perday  %>%
    summarise(
      mean_per_day   = mean(steps_per_day, na.rm = TRUE),
      medium_per_day = median(steps_per_day, na.rm = TRUE)
    )
  #return the plot and the mean and median
  ret <- list("plot" = p, "mean_median" = mean_median)
  ret
}

```

Generate the plots.
Do I need to save these manually??

```{r}
####################################################
#Compare the raw and the padded data

raw <-histo_mean_median ( df.activity, 'Histogram of steps taken per day for the 2 months')
pad <-histo_mean_median ( df.activity.padded, 'Histogram  of padded data - see method - data essentially the same')

# histograms on a panel
grid.arrange(raw$plot, pad$plot, ncol=1, nrow=2)

# the means and median
raw$mean_median
pad$mean_median

```

### What is the average daily activity pattern?


```{r}

```




### Are there differences in activity patterns between weekdays and weekends?


```{r}

```

